Adaptation Writ Large:\\ An Essay in Memory of John H. Holland

### 2015-09-16 paper.tex
Trying again to change my email address
email change
Fixed a few errors in my section

\title{Adaptation Writ Large:\\ An essay in memory of John H. Holland}

%\title{A Common Framework for Defense across Complex Systems}

%\title{Complex Defense: Towards a Theory of Security across Complex Systems}

% Listed authors alphabetically for now.

\numberofauthors{2}
\author 
{\alignauthor
 Stephanie Forrest\\
 \affaddr{University of New Mexico}\\
 \affaddr{Santa Fe Institute}\\
 \email{forrest@cs.unm.edu}
 \alignauthor
Melanie Mitchell \\
 \affaddr{Portland State University}\\
 \affaddr{Santa Fe Institute}\\
 \email{mm@cs.pdx.edu}
}

@@

Prof. John H. Holland passed away recently in Ann Arbor, MI, where he had been on the Univ. of Michigan faculty for XXXX years.  John, as he was known universally to his colleagues and students, 
leaves behind a long legacy of intellectual achievements that is in some ways difficult to pin down.   He is best known for the mechanisms he proposed, especially genetic algorithms, but the motivations behind the mechanisms are less well known but probably more important in the long run.  These mechanisms did not always implement his theoretical framworks perfectly, and in many cases too much effort was spent trying to rescue the mechanisms and not enough in understanding the underlying rationale.  In this short essay, we aim to remind people of some of the key themes of his work, discuss the role they have played in computer science, and highlight especially the ideas that we think remain relevant to today's research agendas.  Many of the problems that John identified and spent his life studying, such as [FILL IN HERE], are still poorly understood.

At a time, when computer science was crytstallizing into an engineering discipline, John came out of the cybernetics era (dating back to von Neumann, Wiener, Ashby, Turing) with a view of computation as a broad interdisciplinary enterprise: "Computation as an idea is something that goes much deeper than operating systems, programming languages, and the like.  The deep ideas of computation are intimately related to the deep ideas of life and intelligence" [Holland quote: Mitchell book, p. 128].  Thus, he became an early proponent of interdisciplinary approaches to computer science and was an active evangelist of what is now called computational thinking, reaching out psychologists, economists, physicists, linguists, philosophers and pretty much anyone he came in contact with.  As a result, even though he received what was arguably the first Ph.D. in our discipline (1958, XXXX), his contributions are much better known outside of computer science than within.

Here we emphasize four recurrent themes of his work: Leveraging statistics in adaptation, learning and credit assignment, models, and the generic properties of complex adaptive systems.

\section{Leveraging statistics in adaptation}

\section{ Learning and Credit assignment}

>>>>

\title{Adaptation Writ Large:\\ An Essay in Memory of John H. Holland}

\numberofauthors{2}
\author 
{\alignauthor
 Stephanie Forrest\\
 \affaddr{University of New Mexico}\\
 \affaddr{Santa Fe Institute}\\
 \email{forrest@cs.unm.edu}
 \alignauthor
Melanie Mitchell \\
 \affaddr{Portland State University}\\
 \affaddr{Santa Fe Institute}\\
 \email{mm@pdx.edu}
}

@@

Professor John H. Holland passed away recently in Ann Arbor, MI, where
he had been on the University of Michigan faculty for over 50 years.
John, as he was known universally to his colleagues and students,
leaves behind a long legacy of intellectual achievements that is in
some ways difficult to pin down.  He is best known for the mechanisms
he proposed, especially genetic algorithms, but the motivations behind
the mechanisms are less well known but probably more important in the
long run.  These mechanisms did not always implement his theoretical
framworks perfectly, and in many cases too much effort was spent
trying to rescue the mechanisms and not enough in understanding the
underlying rationale.  In this short essay, we aim to remind people of
some of the key themes of his work, discuss the role they have played
in computer science, and highlight especially the ideas that we think
remain relevant to today's research agendas.  Many of the problems
that John identified and spent his life studying, such as [FILL IN
  HERE], are still poorly understood.

At a time, when computer science was crytstallizing into an
engineering discipline, John came out of the cybernetics era (dating
back to von Neumann, Wiener, Ashby, Turing) with a view of computation
as a broad interdisciplinary enterprise.  Thus, he became an early
proponent of interdisciplinary approaches to computer science and was
an active evangelist of what is now called computational thinking,
reaching out psychologists, economists, physicists, linguists,
philosophers and pretty much anyone he came in contact with.  As a
result, even though he received what was arguably the first Ph.D. in
our discipline, his contributions are much better known outside of
computer science than within.

Here we emphasize four recurrent themes of his work: adaptation as
exploitation and exploration, learning and credit assignment, models,
and the generic properties of complex adaptive systems.

\section{Adaptation as Exploitation and Exploration}

Although John Holland is best known for inventing genetic algorithms, his
lifelong motivation was much broader: to understand the
phenomenon of adaptation in all its guises.  Holland's interest in
developing a general theory of adaptation was spurred both by his
early work on computer models of Hebbian learning
\cite{RochesterHolland}.  This interest developed further with
Holland's extensive reading in evolutionary biology, economics, game
theory, and control theory.  Holland recognized that all these fields
in essence concern agents that must use information as it is gained from
uncertain, changing environments in order to continually improve their
performance.  Moreover, Holland recognized that in systems with
``adaptive'' agents of this kind, there is never a state of {\emph
  equilibrium} or a final {\emph optimum} configuration; rather, due
to the environment's ``perpetual novelty'' (as Holland termed it),
adaptation continues forever in an open-ended way.

Underlying Holland's theory of adaptation are three key ideas: (1) In
a population undergoing adaptation, individuals can be decomposed into
{\emph building blocks}---traits that are the ``atoms'' of an
individual's fitness or performance.  (2) Successful adaptation
requires the right tradeoff between {\emph exploitation}, in which
tried-and-true building blocks propagate in a population, and {\emph
  exploration}, in which new building blocks (or recombinations of
existing building blocks) arise and are subject to the test of the
environment. (3) Adaptive evolution (biological and otherwise) is a
population-based stochastic process that results in a nearly optimal
exploitation versus exploration tradeoff.

Holland's early papers (e.g.,
\cite{OutlineLogicalTheory,OptimalAllocation}) and his influential
1975 book {\emph Adaptation in Natural and Artificial Systems}
\cite{ANAS} developed a general, formal setting in which these ideas
could be expressed mathematically.  Moreover, Holland stated and
proved a set of theorems that captured the third idea: that, given certain
assumptions, his model of adaptation, while explicitly modifying
populations of individuals, implicitly searched the space of possible
building blocks in a way that optimized the exploitation and
exploration balance.  (As an earlier part of this work, Holland
separately derived an equation for that optimal balance, via a
so-called ``multi-armed bandit'' model.)

It was this formal setting and resulting theorems that led to the
invention of genetic algorithms, which featured stochastic population-based
search and recombination (or {\emph crossover}) as the most important
operation.  However, it is important to note that
the framework developed in \cite{ANAS} was more general than the
genetic algorithm and was created in order to develop an
interdisciplinary theory of adaptation, one that would inform biology,
say, as much as computer science \cite{ChristiansenFeldmanPaper}.  The
later, successful application of genetic algorithms to real-world
optimization and learning tasks was, for Holland, just icing on the
cake.

\section{Learning Via Credit Assignment and Rule Discovery}
